---
title: "MAF Analysis with maftools - Filtered by COSMIC Genes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load Libraries and MAF Data

```{r}
library(maftools)

maf_file <- "/home/akorosteleva/tutorial/example_2/results/maf_files/merged_8_patients.maf"

# Create example clinical data with survival info for 8 patients
clinical_df <- data.frame(
  Tumor_Sample_Barcode = c(
    "P10_cSCC_WES_SRR11870485", "P2_cSCC_WES_SRR11870469",
    "P3_cSCC_WES_SRR11870471", "P5_cSCC_WES_SRR11870475",
    "P6_cSCC_WES_SRR11870477", "P7_cSCC_WES_SRR11870479",
    "P8_cSCC_WES_SRR11870481", "P9_cSCC_WES_SRR11870483"
  ),
  Overall_Survival_Status = c(1, 1, 0, 1, 0, 1, 1, 0),
  Overall_Survival_Time = c(24, 18, 36, 12, 40, 10, 20, 32),
  stringsAsFactors = FALSE
)

maf_data <- read.maf(maf = maf_file, clinicalData = clinical_df)
```

## 2. Summary and Overview

```{r}
getSampleSummary(maf_data)
getGeneSummary(maf_data)
getFields(maf_data)

plotmafSummary(maf = maf_data, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE)
oncoplot(maf = maf_data, top = 30, legend_height = 6, fontSize = 0.5)
```

## 3. Filter Based on COSMIC Genes

```{r}
cosmic_file <- "/home/akorosteleva/tutorial/example_2/working_folder/Cosmic_CancerGeneCensus_v102_GRCh38.tsv"
cosmic_data <- read.delim(cosmic_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
gene_names <- unique(cosmic_data$GENE_SYMBOL)

maf_data_filtered <- subsetMaf(maf = maf_data, genes = gene_names)
oncoplot(maf = maf_data_filtered, top = 30, legend_height = 6, fontSize = 0.5)
```

## 4. Transitions and Transversions

```{r}
titv = titv(maf = maf_data_filtered, plot = FALSE, useSyn = TRUE)
plotTiTv(res = titv)
```

> It's important to keep in mind that not all of these mutations are drivers or even relevant to this cancer type. Many may have no functional impact. Therefore, additional filtering based on predicted functional effect (e.g., high or moderate impact) and clinical annotation should also be considered.

## 5. Keep Only Pathogenic Variants

```{r}
# Extract the MAF as a data frame
maf_df <- maf_data_filtered@data

# Convert to lowercase and trim whitespace
maf_df$CLIN_SIG_clean <- trimws(tolower(maf_df$CLIN_SIG))

# Define allowed CLIN_SIG values (also lowercase and trimmed)
keep_values <- c(
  "pathogenic,pathogenic/likely_pathogenic,likely_pathogenic",
  "pathogenic,pathogenic/likely_pathogenic",
  "pathogenic",
  "pathogenic,likely_pathogenic",
  "drug_response,pathogenic,likely_pathogenic,not_provided",
  "likely_pathogenic,pathogenic",
  "likely_pathogenic"
)

# Filter
maf_df_strict <- maf_df[maf_df$CLIN_SIG_clean %in% keep_values, ]

maf_data_filtered_strict <- read.maf(
  maf = maf_df_strict,
  clinicalData = clinical_df
)

oncoplot(maf_data_filtered_strict, top=30)
```

## 6. Suggested Next Steps

Please read  
[maftools vignette](https://bioconductor.org/packages/devel/bioc/vignettes/maftools/inst/doc/maftools.html)  
and explore other functions such as:  
- Survival Analysis  
- Detecting Cancer Driver Genes  
- Lollipop Plots  
- Drug-Gene Interactions  
- Oncogenic Signaling Pathways
